State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-05-08 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
periodic commit. almost completed rough draft of discussion

\section{Discussion}\label{DiscussionSection}
This paper presents a novel framework for model reduction to make estimation of neural field practical. This systematic model reduction allows for the choice a complexity. The level of detail of the field is related to the number and width of the field basis functions, thereby providing a flexible approach where detail and computational complexity can be tradeoff. For certain application it might be sufficient to capture only low (spatial) frequency dynamics.

The frequency analysis and model selection tools presented in this paper specifies the sensor and field basis function arrangement required to form a state-space model to capture and model dominant cortical dynamics. These tools can be used to better design experiments and recording techniques to advance model based neural parameter estimation and avoid spatial aliasing. In addition, the assumption homogeneity can be relaxed where heterogeneity can be model by careful placement of field basis functions. For example, if cortical regions have a greater density of neural populations, then more basis functions can be used in these areas. In this scenario, model selection and basis function positioning can be achieved using instantaneous frequency analysis of the the neural field or alternative model selection tools based on information criteria.

To the authors best knowledge, this is the first paper to propose an estimation framework for intracortical connectivity from electrophysiological data. This is important as the connectivity structure at this scale can not be estimated using diffusion tensor imaging due to the fine spatial scale and the high degree of crossing of fibres\omg{ref}. In addition, other non-model based techniques for estimating functional connectivity such as Granger causality \cite{Hesse2003}, the direct transfer function \cite{Kaminski1991}, and partial directed coherence \cite{Sameshima1999} are restricted to the spatial scales of the recording electrodes. The model based technique proposed in this paper estimates the connectivity structure independently of the recording electrode geometry. An advantage of these auto-regressive (AR) based methods is that they allow for anisotropic connectivity, which is present in cortico-cortical fibres. The AR methods are complementary to the proposed model based framework and by combining these methods a more realistic model can be estimated. A recent theoretical study~\cite{Jirsa2009} has demonstrated the importance of this multi-scale structure in generating the characteristic rhythms of the brain.

Long range connectivity can be estimated using DTI \cite{Knock2009}, however this short range connectivity can not estimated using DTI

Generating patient specific neural field models has the potential to have significant impact on several areas of neuroscientific research and clinical neurology. Specifically, the development of a patient-specific state-space model will allow for the application of systems theoretical techniques from engineering to be applied to neuro-dynamics. For example, state tracking could be used in brain-computer interface or epileptic seizure prediction applications, where specific regions of the state-space may indicate intent of movement or imminent seizures. In addition, a patient specific state-space model will allow application of control engineering techniques to robustly prevent the neuro-dynamics entering pathological regions of state-space using electrical stimulation.

@@

This estimation framework also has theoretical implications. This estimation framework has the potential to allow testing hypotheses that have been generated in theoretical studies and to validate neural field models. For example, it has been hypothesised that so called bump solutions are a possible mechanism for short term memory formation~\cite{Coombes2005}. Using this framework, we can reconstruct a neural field from data, and estimate parameters and check if they correspond the the theoretically derived parameters where these bump solutions can exist. Therefore, this framework has the potential to validate models and test theoretically derived hypotheses. Associating regions of state-space and parameter sets to cognitive can potentially validate theoretical findings. For example, sleep, wake, memory,

Here we assumed knowledge of the width of the kernel basis functions. If a future work we plan to extend this work to unknown widths.

In future work, we aim to show how heterogeneous can be modelled within this formulation. We will still assume a isotropic connectivity kernel however heterogeneities can be modelled by placing basis function c

\begin{itemize}
	\item if we don't know the frequency analysis or the system is marginally sampled or under-sampled, the best we can do is have the basis functions the same as the sensor kernel. \wtf{why wouldn't it be better to have, say, twice as many basis functions as we have sensors, at half the width?} 
	\item alternatives to frequency analysis are model selection... \wtf{This frequency analysis \emph{is} model selection. Maybe what's meant here is that other model selection approaches could be applied? Some sort of IC optimisation?}

	\item The size of the dentritic arbour is consistent in the cortex. \omg{?}

	\item The need for activity to perform estimation. \omg{To Learn You Must Perturb!} 
	\item Recent studies using neural field models have demonstrated the theoretical implications of specific connectivity structure of the cortical columns in neural field models, where the connectivity kernel governs the bifurcation points of the system \cite{Hutt2005} and types of oscillations that can be generated \cite{Schmidt2009}. This implies that if we could estimate the connectivity structure for an individual, then we could capture patient specific neurodynamics that lead to various oscillatory states.
\end{itemize}

>>>>

\section{Discussion}\label{DiscussionSection}
This paper presents a novel framework for model reduction to facilitate practical state and parameter estimation of neural fields. The level of detail of the estimated field is related to the basis functions decomposition, thereby providing a flexible approach where a trade-off can be made between physiological detail and computational complexity.

The frequency analysis and model selection tools presented in this paper specify the sensor and field basis function arrangement required to form a state-space model to capture and model dominant cortical dynamics. These tools can be used to better design electrophysiological recording techniques to advance model-based neural parameter estimation to avoid spatial aliasing. In addition, the assumption homogeneity can be relaxed, where heterogeneity can be modelled by careful placement of field basis functions. For example, if cortical regions have a greater density of neural populations, then more basis functions can be used in these areas. In this scenario, model selection and basis function positioning can be achieved using instantaneous frequency analysis of the the neural field or alternative model selection tools based on information criteria.

To the authors best knowledge, this is the first paper to propose an estimation framework for intracortical connectivity from electrophysiological data. The importance of this can be illustrated by considering the connectivity structure at this scale can not be estimated using diffusion tensor imaging due to the isotropic diffusion or water in grey matter, the fine spatial scale and the high degree of crossing of fibres~\cite{Assaf2008}. In addition, other non-model based techniques for estimating functional connectivity such as Granger causality \cite{Hesse2003}, the direct transfer function \cite{Kaminski1991}, and partial directed coherence \cite{Sameshima1999} are restricted to the spatial scales of the recording electrodes. The model based technique proposed in this paper estimates the connectivity structure independently of the recording electrode geometry. An advantage of the auto-regressive (AR) based methods is that they allow for anisotropic connectivity, which is present in cortico-cortical fibres. The AR and the proposed model-based methods are complementary, where they estimating connectivity at different spatial scales. Therefore, by combining these methods a more realistic model can be estimated. A recent theoretical study~\cite{Jirsa2009} has demonstrated the importance of this multi-scale structure in generating the characteristic rhythms of the brain.

Generating patient specific neural field models has the potential to have significant impact on several areas of neuroscientific research and clinical neurology. Specifically, the development of a patient-specific state-space model will allow for the application of systems theoretical techniques from engineering to be applied to neuro-dynamics. For example, state tracking could be used in brain-computer interface or epileptic seizure prediction applications, where specific regions of the state-space may indicate intent of movement or imminent seizures. Recent studies using neural field models have demonstrated the theoretical implications of specific connectivity structure of the cortical columns in neural field models, where the connectivity kernel governs where a Turing-Hoft bifurcation occurs~\cite{Hutt2005}, and types of oscillations that can be generated~\cite{Schmidt2009}. This implies that if we could estimate the connectivity structure for an individual, then we could capture patient specific neurodynamics that lead to various oscillatory states. In addition, a patient specific state-space model will allow application of control engineering techniques to robustly prevent the neuro-dynamics entering pathological regions of state-space using electrical stimulation.

@@

This estimation framework also has theoretical implications. This estimation framework has the potential to allow testing hypotheses that have been generated in theoretical studies and to validate neural field models. For example, it has been hypothesised that so called bump solutions are a possible mechanism for short term memory formation~\cite{Coombes2005}. Using the proposed model-based framework, we can reconstruct a neural field from data, and estimate parameters and check if they correspond the the theoretically derived parameters where these bump solutions can exist. In addition exploration of parameters and the state-space can be explored during other cognitive tasks and states of arousal such as sleep and anaesthesia.

\begin{itemize}
	\item Here we assumed knowledge of the width of the kernel basis functions. If a future work we plan to extend this work to unknown widths. 
	\item The size of the dentritic arbour is consistent in the cortex. This implies global isotropy. Heterogeneity is less consistent. 
	\item The need for activity to perform estimation. \omg{To Learn You Must Perturb!} 
\end{itemize}
